---
title: "ICE Annina 10x sequencing analysis - Analysis of human data"
author: "Vladyslav Korobeynyk at Jessberger/Robinsonlab, UZH"
date: last-modified
format:
  html:
    toc: true
    css: swt.css
    linkcolor: '#E5005C'
    df-print: kable
    embed-resources: true
    code-fold: true
    code-block-bg: '#E6EEF0'
    code-block-border-left: '#3F698A'
params:
  metacell_dir: "/home/vkorob/data/aging_human_organoid_comparisson/Human_analysis/human_analysis_output"
  dea_output_dir: "/home/vkorob/data/aging_human_organoid_comparisson/Human_analysis/human_analysis_output/DEA_results"
  gsea_output_dir: "/home/vkorob/data/aging_human_organoid_comparisson/Human_analysis/human_analysis_output/GSEA_analysis"
  helper_functions_path: "/home/vkorob/data/aging_human_organoid_comparisson/helper_functions.R"
  qc_pipeline_functions_path: "/home/vkorob/data/aging_human_organoid_comparisson/general_qc_sc_pipeline_R_function.R"
  n_top_genes_per_direction: 2
  n_gsea_pathways: 150
---

::: callout-note
**What this script does:** loads the metacell-level `SingleCellExperiment` objects
(Inhibitory and Excitatory neurons) produced by `01_human_Processing.qmd`, then for
each cell type — and for the two combined — checks how much of the variance in
expression is explained by Age and other covariates (PMI, Sex, Diagnosis,
Sample_Source, Individual_ID), runs a metacell-level DESeq2 differential expression
analysis of Age (controlling for Diagnosis, Sample_Source, Sex, and Celltype where
relevant), and visualizes expression of the top age-associated genes. The resulting
DEA marker tables are then used to run GSEA against GO Biological Process terms,
producing reduced/clustered GO-term treemaps and saved marker/pathway tables that
feed into the cross-dataset (organoid vs. human) comparison later in the project.
:::

# Setup

```{r setup, tags=c(), warning=FALSE, message=FALSE}
options(future.globals.maxSize = 8000 * 1024^2)
suppressMessages({
  library(Seurat)
  library(harmony)
  library(scater)
  library(caret)
  library(dplyr)
  library(parallel)
  library(ggpubr)
  library(ggplot2)
  library(Matrix)
  library(stringr)
  library(magrittr)
  library(EnhancedVolcano)
  library(glmnet)
  library(genekitr)
  library(geneset)
  library(DESeq2)
  library(pheatmap)
  library(tidyr)
  library(ggVennDiagram)
  library(sva)
  library(rrvgo)
})
source(params$helper_functions_path)
source(params$qc_pipeline_functions_path)
dir.create(params$dea_output_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(params$gsea_output_dir, recursive = TRUE, showWarnings = FALSE)

gs = geneset::getGO(org = "human", ont = "bp")
```

# Load data

```{r load-metacells, warning=FALSE, message=FALSE, results = 'hide'}
metacell_sce_IN = readRDS(file.path(params$metacell_dir, "metacell_sce_IN_human.RDS"))
metacell_sce_EN = readRDS(file.path(params$metacell_dir, "metacell_sce_EN_human.RDS"))
```

# Helper functions for this script

The DEA + diagnostic-plot workflow functions

```{r dea-plot-functions, warning=FALSE, message=FALSE}
#' Run variance-explained diagnostics + metacell-level DESeq2 DEA, save markers, plot MA
#'
#' @param sce SingleCellExperiment of metacells
#' @param design_formula DESeq2 design formula
#' @param contrast passed to metacell_DESeq2_DEA
#' @param organism "human" or "organoid", passed to metacell_DESeq2_DEA
#' @param markers_csv_path where to write the DEA markers table
#' @param variance_df optional data.frame of covariates for getVarianceExplained/
#'   plotExplanatoryVariables; skipped if NULL (not run for the merged IN|EN analysis,
#'   matching the original design)
#' @return list(markers = ..., DESeq2_lst = ...)
run_metacell_dea = function(sce, design_formula, contrast, organism,
                             markers_csv_path, variance_df = NULL) {

  if (!is.null(variance_df)) {
    getVarianceExplained(sce, variables = variance_df) %>%
      plotExplanatoryVariables() %>%
      print()
  }

  sce$Age = as.numeric(sce$Age)

  DESeq2_lst = metacell_DESeq2_DEA(
    sce = sce, phenotype_column = "Age", dataset_column = "Individual_ID",
    design_formula = design_formula, contrast = contrast, organism = organism
  )

  # Sample correlation heatmap, annotated by Age/diagnosis/sample source
  print(DESeq2_lst$heatmap)
  
  markers = DESeq2_lst$markers %>% filter(if_all(everything(), ~ !is.na(.)))
  write.csv(markers, markers_csv_path)

  DESeq2::plotMA(DESeq2_lst$res)

  list(markers = markers, DESeq2_lst = DESeq2_lst)
}

#' Plot VST-normalized expression vs Age for the top up/down DEA genes,
#' colored by each covariate in color_by_vars (one plot per covariate)
plot_top_age_genes = function(markers, DESeq2_lst, color_by_vars, n_top_genes) {

  top_down = markers %>% arrange(avg_log2FC) %>% dplyr::slice(1:n_top_genes) %>% pull(gene)
  top_up   = markers %>% arrange(desc(avg_log2FC)) %>% dplyr::slice(1:n_top_genes) %>% pull(gene)
  genes = c(top_down, top_up)

  expr = assay(DESeq2_lst$vst_transformed_deseq2)  # genes x samples
  dds  = DESeq2_lst$dds

  df = expr[genes, , drop = FALSE] %>%
    as.data.frame() %>%
    tibble::rownames_to_column("gene") %>%
    pivot_longer(-gene, names_to = "sample", values_to = "expression") %>%
    left_join(
      as.data.frame(colData(dds)) %>% tibble::rownames_to_column("sample"),
      by = "sample"
    )

  for (covariate in color_by_vars) {
    p = ggplot(df, aes(x = Age, y = expression, color = .data[[covariate]])) +
      geom_jitter(width = 0.2, size = 1.5, alpha = 0.6) +
      facet_wrap(~ gene, scales = "free_y") +
      theme_minimal() +
      labs(y = "VST normalized expression") +
      ggtitle(paste0(n_top_genes, " down / ", n_top_genes,
                      " up genes changing with age, colored by ", covariate))
    print(p)
  }
}

#' Run GSEA on a saved DEA markers table, save results, generate reduced-GO treemaps
run_gsea_workflow = function(markers_csv_path, gs, n_pathways, title_label,
                              gsea_rds_path, neg_txt_path, pos_txt_path) {

  markers = read.csv(markers_csv_path, row.names = 1)
  gsea = gsea_analysis(markers = markers, cluster = "trending clusters", gs = gs,
                        n_pathways = n_pathways, plot_type = "bar")

  if (is.list(gsea)) {
    saveRDS(gsea, gsea_rds_path)
    lst = plot_ReducedGOterms(gsea_df = gsea$gsea_df, title_label = title_label)
    write.table(lst$neg, neg_txt_path, row.names = FALSE, quote = FALSE)
    write.table(lst$pos, pos_txt_path, row.names = FALSE, quote = FALSE)
  }

  gsea
}
```

# MDS plot / Var explained / DEA

**Here we are trying to understand if Age is a variable that can explain the variability in the data**

::: panel-tabset
## IN

```{r MDS_DEA_iN, warning=FALSE, message=FALSE, results = 'hide', fig.height=8, fig.width=8}
variance_df_IN = data.frame(
  Age = metacell_sce_IN$Age,
  PMI = metacell_sce_IN$PMI,
  Sex = metacell_sce_IN$Sex,
  Diagnosis = metacell_sce_IN$Diagnosis,
  Sample_Source = metacell_sce_IN$Sample_Source,
  Individual_ID = metacell_sce_IN$Individual_ID
)

dea_IN = run_metacell_dea(
  sce = metacell_sce_IN,
  design_formula = as.formula("~ Diagnosis + Sample_Source + Sex + Age"),
  contrast = "Age",
  organism = "human",
  markers_csv_path = file.path(params$dea_output_dir, "DEA_metacell_iN_old_vs_young.csv"),
  variance_df = variance_df_IN
)

plot_top_age_genes(
  markers = dea_IN$markers, DESeq2_lst = dea_IN$DESeq2_lst,
  color_by_vars = c("Sample_Source", "Sex", "Diagnosis"),
  n_top_genes = params$n_top_genes_per_direction
)
```

## EN

```{r MDS_DEA_exN, warning=FALSE, message=FALSE, results = 'hide', fig.height=10, fig.width=10}
variance_df_EN = data.frame(
  Age = metacell_sce_EN$Age,
  PMI = metacell_sce_EN$PMI,
  Sex = metacell_sce_EN$Sex,
  Diagnosis = metacell_sce_EN$Diagnosis,
  Sample_Source = metacell_sce_EN$Sample_Source,
  Individual_ID = metacell_sce_EN$Individual_ID
)

dea_EN = run_metacell_dea(
  sce = metacell_sce_EN,
  design_formula = as.formula("~ Diagnosis + Sample_Source + Sex + Age"),
  contrast = "Age",
  organism = "human",
  markers_csv_path = file.path(params$dea_output_dir, "DEA_metacell_exN_old_vs_young.csv"),
  variance_df = variance_df_EN
)

plot_top_age_genes(
  markers = dea_EN$markers, DESeq2_lst = dea_EN$DESeq2_lst,
  color_by_vars = c("Sample_Source", "Sex", "Diagnosis"),
  n_top_genes = params$n_top_genes_per_direction
)
```

## IN | EN

Merge both cell types and perform similar analysis as above

```{r MDS_DEA_iN_exN, warning=FALSE, message=FALSE, results = 'hide', fig.height=8, fig.width=8}
# Prefix cell IDs so IN/EN metacells don't collide when merged, and restrict to shared genes/colData
sce_IN_for_merge = metacell_sce_IN
sce_EN_for_merge = metacell_sce_EN
colnames(sce_IN_for_merge) = paste0("IN_", colnames(sce_IN_for_merge))
colnames(sce_EN_for_merge) = paste0("EN_", colnames(sce_EN_for_merge))

common_genes = intersect(rownames(sce_IN_for_merge), rownames(sce_EN_for_merge))
sce_IN_for_merge = sce_IN_for_merge[common_genes, ]
sce_EN_for_merge = sce_EN_for_merge[common_genes, ]

common_cols = intersect(names(colData(sce_IN_for_merge)), names(colData(sce_EN_for_merge)))
colData(sce_IN_for_merge) = colData(sce_IN_for_merge)[, common_cols, drop = FALSE]
colData(sce_EN_for_merge) = colData(sce_EN_for_merge)[, common_cols, drop = FALSE]

merged_metacells_sce = cbind(sce_IN_for_merge, sce_EN_for_merge)

variance_df_merged = data.frame(
  Age = merged_metacells_sce$Age,
  PMI = merged_metacells_sce$PMI,
  Sex = merged_metacells_sce$Sex,
  Diagnosis = merged_metacells_sce$Diagnosis,
  Sample_Source = merged_metacells_sce$Sample_Source,
  Individual_ID = merged_metacells_sce$Individual_ID,
  CellType = merged_metacells_sce$CellType
)

dea_merged = run_metacell_dea(
  sce = merged_metacells_sce,
  design_formula = as.formula("~ Diagnosis + Sample_Source + Sex + CellType + Age"),
  contrast = "Age",
  organism = "human",
  markers_csv_path = file.path(params$dea_output_dir, "DEA_metacell_neurons_old_vs_young.csv"),
  variance_df = variance_df_merged
)
```
:::

# Patient GSEA

::: panel-tabset
## IN

```{r GSEA_IN_metacell, fig.height=10, fig.width=10}
gsea_IN = run_gsea_workflow(
  markers_csv_path = file.path(params$dea_output_dir, "DEA_metacell_iN_old_vs_young.csv"),
  gs = gs,
  n_pathways = params$n_gsea_pathways,
  title_label = "human metacell IN cells across age",
  gsea_rds_path = file.path(params$gsea_output_dir, "GOterms_metacell_old_vs_young_iN_all_old.RDS"),
  neg_txt_path = file.path(params$gsea_output_dir, "GOterms_metacell_old_vs_young_iN_downregulated_old.txt"),
  pos_txt_path = file.path(params$gsea_output_dir, "GOterms_metacell_old_vs_young_iN_upregulated_old.txt")
)
```

## EN

```{r GSEA_EN_metacell, fig.height=10, fig.width=10}
gsea_EN = run_gsea_workflow(
  markers_csv_path = file.path(params$dea_output_dir, "DEA_metacell_exN_old_vs_young.csv"),
  gs = gs,
  n_pathways = params$n_gsea_pathways,
  title_label = "human metacell EN cells across age",
  gsea_rds_path = file.path(params$gsea_output_dir, "GOterms_metacell_old_vs_young_exN_all_old.RDS"),
  neg_txt_path = file.path(params$gsea_output_dir, "GOterms_metacell_old_vs_young_exN_downregulated_old.txt"),
  pos_txt_path = file.path(params$gsea_output_dir, "GOterms_metacell_old_vs_young_exN_upregulated_old.txt")
)
```

## IN | EN

```{r GSEA_INEN_metacell, fig.height=10, fig.width=10}
# NOTE: title fixed to reflect IN|EN (original script mistakenly reused the "EN" title here)
gsea_merged = run_gsea_workflow(
  markers_csv_path = file.path(params$dea_output_dir, "DEA_metacell_neurons_old_vs_young.csv"),
  gs = gs,
  n_pathways = params$n_gsea_pathways,
  title_label = "human metacell IN|EN cells across age",
  gsea_rds_path = file.path(params$gsea_output_dir, "GOterms_metacell_old_vs_young_neurons.RDS"),
  neg_txt_path = file.path(params$gsea_output_dir, "GOterms_metacell_old_vs_young_neurons_downregulated_old.txt"),
  pos_txt_path = file.path(params$gsea_output_dir, "GOterms_metacell_old_vs_young_neurons_upregulated_old.txt")
)
```
:::

# sessioninfo

```{r load_data}
sessionInfo()
```
