---
title: "ICE Annina 10x sequencing analysis - generate metacells for DEA/GSEA"
author: "Vladyslav Korobeynyk at Jessberger/Robinsonlab, UZH"
date: last-modified
format:
  html:
    toc: true
    css: swt.css
    linkcolor: '#E5005C'
    df-print: kable
    embed-resources: true
    code-fold: true
    code-block-bg: '#E6EEF0'
    code-block-border-left: '#3F698A'
params:
  object_storage_path: "/home/vkorob/data/aging_human_organoid_comparisson/Organoid_analysis/organoid_analysis_output"
  gsea_results_path: "/home/vkorob/data/aging_human_organoid_comparisson/Organoid_analysis/organoid_analysis_output/GSEA_results/metacells"
  dea_results_path: "/home/vkorob/data/aging_human_organoid_comparisson/Organoid_analysis/organoid_analysis_output/DEA_results/metacells"
  helper_functions_path: "/home/vkorob/data/aging_human_organoid_comparisson/helper_functions.R"
  qc_pipeline_functions_path: "/home/vkorob/data/aging_human_organoid_comparisson/general_qc_sc_pipeline_R_function.R"
  amount_of_cells_to_aggregate: 15
  n_metacells_generate: 20
  gsea_n_pathways: 150
  dea_padj_threshold: 0.05
---

::: callout-note
**What this script does:** loads the integrated/annotated organoid Seurat object from
`01_organoid_QC_processing_integration.qmd`, generates metacells (bootstrapped
pseudobulk profiles) per celltype for each dataset grouping (COp1 10w, COp1 4m, COp2
4dpi+4wpi pooled, and COp2 4dpi/4wpi separately for IN/EN), runs ICE-vs-wt DESeq2 DEA
and GSEA (GO Biological Process) on each, and saves the resulting metacell objects
and DEA/GSEA tables for use in later scripts (organoid single-cell analysis, and the
human/organoid ML prediction scripts). It also runs a few "all datasets combined" DEA
and GSEA analyses to look at the ICE-vs-wt signature across the whole organoid cohort.
:::

# libraries

```{r libs, warning=FALSE, message=FALSE}
options(scipen = 1, digits = 3) 
set.seed(1)

suppressMessages({
  library(dplyr)
  library(Seurat)
  library(edgeR)
  library(harmony)
  library(stringr)
  library(stringi)
  library(genekitr)
  library(ggplot2)
  library(geneset)
  library(scater)
  library(magrittr)
  library(SingleCellExperiment)
  library(DT)
  library(EnhancedVolcano)
  library(cluster)
  library(DESeq2)
  library(pheatmap)
  library(rrvgo)
  library(tidyr)
  library(openxlsx)
  library(ggpubr)
  library(patchwork)
})

# Load local pipeline utilities
source(params$qc_pipeline_functions_path)
source(params$helper_functions_path)

# Output directory path configurations
object_storage_path = params$object_storage_path
gsea_results_path = params$gsea_results_path
DEA_results_path = params$dea_results_path

dir.create(DEA_results_path, recursive = TRUE, showWarnings = FALSE)
dir.create(gsea_results_path, recursive = TRUE, showWarnings = FALSE)

gs = geneset::getGO(org = "human", ont = "bp")
```

# Load object

```{r load object, warning=FALSE, message=FALSE}
harmonized_seurat = readRDS(file.path(object_storage_path , "SO_integratedHarmony_alldatasetsTogether_organoids.RDS"))
harmonized_seurat_COp1_10w = subset(harmonized_seurat, Protocol_maturation_stage == "COp1_10w")
harmonized_seurat_COp1_4m = subset(harmonized_seurat, Protocol_maturation_stage == "COp1_4m")
harmonized_seurat_COp2_4dpi_4wpi = subset(harmonized_seurat, Protocol == "COp2")
```

# Helper functions for this script

```{r metacell-helper-functions, warning=FALSE, message=FALSE}
#' Generate metacells + per-celltype DESeq2 DEA (ICE vs wt) for a Seurat subset
#' @param SO Seurat object already subset to the dataset/timepoint grouping of interest
#' @param celltypes character vector of Celltype values to loop over
#' @return named list (by celltype) of list(sce = , markers = , DESeq2_lst = )
generate_metacells_dea_per_celltype = function(SO, celltypes, amount_of_cells_to_aggregate, n_metacells_generate) {
  lapply(celltypes, function(ct) {
    sce = create_metacells(
      SO = subset(SO, Celltype == ct),
      amountOfCells_toAggregate = amount_of_cells_to_aggregate,
      N_metacells_generate = n_metacells_generate
    )
    sce$phenotype = as.factor(sce$phenotype)

    DESeq2_lst = metacell_DESeq2_DEA(
      sce = sce, phenotype_column = "phenotype", dataset_column = "dataset",
      design_formula = as.formula("~ phenotype"), title_corHeatmap = paste0("metacell correlation ", ct),
      contrast = c("phenotype", "ICE", "wt"), organism = "organoid"
    )

    markers = DESeq2_lst$markers %>% na.omit()
    list(sce = sce, markers = markers, DESeq2_lst = DESeq2_lst)
  }) %>% setNames(celltypes)
}

#' Run GSEA on each celltype's DEA markers, and summarize the (relative) number of
#' significant GO terms found per celltype
#' @return list(out = per-celltype GSEA results, goterm_summary = data.frame for plotting)
summarize_gsea_per_celltype = function(metacell_list, gs, n_pathways) {
  out = lapply(metacell_list, function(x) {
    sign_markers = x$markers %>% filter(padj < params$dea_padj_threshold)
    gsea = gsea_analysis(markers = x$markers, cluster = "trending clusters", gs = gs,
                          n_pathways = n_pathways, plot_type = "none")
    list(sign_markers = sign_markers, n_sign_markers = nrow(sign_markers), gsea = gsea)
  })

  goterm_summary = lapply(out, function(x) {
    if (is.list(x$gsea)) nrow(x$gsea$gsea_df)
  }) %>%
    Filter(Negate(is.null), .) %>%
    as.data.frame() %>%
    t() %>%
    as.data.frame() %>%
    tibble::rownames_to_column("celltype") %>%
    dplyr::rename(n = V1) %>%
    mutate(relative_n = n / sum(n))

  list(out = out, goterm_summary = goterm_summary)
}

#' Show PCA/MA diagnostics for a metacell DESeq2 result, save the markers table, and
#' plot expression of the top genes by padj across phenotype
plot_dea_tab = function(DESeq2_lst, output_csv_path, n_top_genes = 4) {
  print(DESeq2_lst$plt_phenotype)
  print(DESeq2_lst$plt_dataset)

  markers = DESeq2_lst$markers %>% na.omit()
  write.csv(markers, output_csv_path)

  DESeq2::plotMA(DESeq2_lst$res)
  with(markers %>% filter(padj < params$dea_padj_threshold) %>% head(10), {
    text(baseMean, avg_log2FC, labels = gene, pos = 3, cex = 1)
    segments(baseMean, avg_log2FC, baseMean, avg_log2FC, col = "gray60")
  })

  # After a VST transformation, values are on a scale where variance is stabilized and no
  # longer mean-dependent, so they're comparable across genes regardless of original variance.
  vst = DESeq2_lst$vst_transformed_deseq2
  dds = DESeq2_lst$dds
  expr = assay(vst)

  genes = markers %>% arrange(padj) %>% dplyr::slice(1:n_top_genes) %>% pull(gene)

  df = expr[genes, , drop = FALSE] %>%
    as.data.frame() %>%
    tibble::rownames_to_column("gene") %>%
    pivot_longer(-gene, names_to = "sample", values_to = "expression") %>%
    left_join(as.data.frame(colData(dds)) %>% tibble::rownames_to_column("sample"), by = "sample")

  print(
    ggplot(df, aes(x = phenotype, y = expression)) +
      geom_boxplot() +
      geom_jitter(width = 0.2, size = 1.5, alpha = 0.6) +
      facet_wrap(~ gene, scales = "free_y") +
      labs(y = "VST normalized expression")
  )

  markers
}

#' Run GSEA on a saved DEA markers CSV, write up/down GO term tables, plot the reduced-GO treemap
run_gsea_tab = function(markers_csv_path, gs, n_pathways, up_txt_path, down_txt_path, title_label) {
  markers = read.csv(markers_csv_path, row.names = 1)
  gsea = gsea_analysis(markers = markers, cluster = "trending clusters", gs = gs,
                        n_pathways = n_pathways, plot_type = "none")

  write.table(gsea$gsea_df %>% filter(NES > 0) %>% select(hsapiens_BP_ID, p.adjust),
              up_txt_path, row.names = FALSE, quote = FALSE)
  write.table(gsea$gsea_df %>% filter(NES < 0) %>% select(hsapiens_BP_ID, p.adjust),
              down_txt_path, row.names = FALSE, quote = FALSE)

  plot_ReducedGOterms(gsea_df = gsea$gsea_df, title_label = title_label)
}

#' Generate metacells + DEA for a single celltype restricted to one Protocol timepoint
#' (used for the COp2 4dpi/4wpi IN/EN objects later for ML prediction).
generate_metacell_by_timepoint = function(SO, celltypes, timepoint, amount_of_cells_to_aggregate,
                                           n_metacells_generate, dea_csv_prefix) {
  lapply(celltypes, function(ct) {
    cells_keep = colnames(SO)[SO$Celltype == ct & SO$timepoint == timepoint]

    sce = create_metacells(
      SO = subset(SO, cells = cells_keep),
      amountOfCells_toAggregate = amount_of_cells_to_aggregate,
      N_metacells_generate = n_metacells_generate
    )
    sce$phenotype = as.factor(sce$phenotype)

    DESeq2_lst = metacell_DESeq2_DEA(
      sce = sce, phenotype_column = "phenotype", dataset_column = "dataset",
      design_formula = as.formula("~ phenotype"), contrast = c("phenotype", "ICE", "wt"),
      organism = "organoid"
    )

    markers = DESeq2_lst$markers %>% na.omit()
    write.csv(markers, paste0(dea_csv_prefix, "_", ct, "_COp2_", timepoint, "_organoids.csv"))

    list(sce = sce, sampleCor_heatmap = DESeq2_lst$heatmap)
  }) %>% setNames(celltypes)
}

#' Restrict a list of SingleCellExperiments to their shared genes and cbind them together
merge_sces_common_genes = function(sce_list) {
  common_genes = Reduce(intersect, lapply(sce_list, rownames))
  message("Number of shared genes across ", length(sce_list), " objects: ", length(common_genes))
  subsetted = lapply(sce_list, function(sce) sce[common_genes, ])
  do.call(cbind, subsetted)
}

#' Generate metacells for one celltype, pooling all datasets with a 4dpi/4wpi timepoint
#' annotation together (regardless of protocol/maturation stage)
generate_metacell_celltype_timepointComparison = function(SO, celltype, amount_of_cells_to_aggregate, n_metacells_generate) {
  cells_keep = colnames(SO)[SO$Celltype == celltype & SO$timepoint %in% c("4dpi", "4wpi")]
  sce = create_metacells(
    SO = subset(SO, cells = cells_keep),
    amountOfCells_toAggregate = amount_of_cells_to_aggregate,
    N_metacells_generate = n_metacells_generate
  )
  sce$phenotype = as.factor(sce$phenotype)
  sce$timepoint = factor(sce$timepoint, levels = c("4dpi", "4wpi")) # 4dpi = reference
  sce
}

#' ICE vs wt effect specifically at one timepoint (Wald). Positive avg_log2FC = higher in ICE.
get_phenotype_effect_at_timepoint = function(sce, reference_timepoint, design_formula) {
  sce$timepoint = relevel(as.factor(sce$timepoint), ref = reference_timepoint)
  dds = DESeqDataSetFromMatrix(counts(sce), colData = colData(sce), design = design_formula)
  dds = DESeq(dds)
  results(dds, contrast = c("phenotype", "ICE", "wt")) %>%
    as.data.frame() %>%
    dplyr::rename("avg_log2FC" = "log2FoldChange") %>%
    mutate(gene = rownames(.))
}

#' LRT: does the phenotype (ICE vs wt) effect differ by timepoint (4dpi vs 4wpi)?
run_timepoint_phenotype_interaction_dea = function(sce, design_full, design_reduced) {
  dds_full = DESeqDataSetFromMatrix(counts(sce), colData = colData(sce), design = design_full)
  dds_full = DESeq(dds_full, test = "LRT", reduced = design_reduced)

  # NOTE: for a multi-coefficient LRT, avg_log2FC does not reflect a clean "higher in X vs Y" direction - only padj
  # from this table should be interpreted (does the phenotype effect differ by
  # timepoint).
  markers = results(dds_full) %>%
    as.data.frame() %>%
    dplyr::rename("avg_log2FC" = "log2FoldChange") %>%
    mutate(gene = rownames(.)) %>%
    filter(!is.na(padj))

  list(dds = dds_full, markers = markers)
}
```

# Generate metacells + DEA | COp1 10w | per Celltype

::: callout-tip
Why Metacells?
Single-cell RNA-seq data is notoriously sparse and prone to high dropout rates. Aggregating transcriptomic profiles into metacells allows us to perform more sensitive, un-biased pseudobulk DEA testing while preserving sample-level replicate fidelity.
:::

```{r ,warning=FALSE, message=FALSE, results = 'hide'}
#| fig-width: 12
#| fig-height: 10

celltypes = unique(harmonized_seurat_COp1_10w$Celltype)

sce_metacells_alldatasets_perCT_10w = generate_metacells_dea_per_celltype(
  harmonized_seurat_COp1_10w, celltypes, params$amount_of_cells_to_aggregate, params$n_metacells_generate
)

saveRDS(sce_metacells_alldatasets_perCT_10w , file.path(object_storage_path , "metacell_sce_allCelltypes_COp1_10w.RDS"))

markers = sce_metacells_alldatasets_perCT_10w$exN$markers
EnhancedVolcano(markers,
              lab = rownames(markers),
              x = 'avg_log2FC',
              y = 'padj',
              title = paste0('Metacell COp1_10w ICE vs wt - ', "exN"),
              pCutoff = 0.05,
              FCcutoff = 1,
              pointSize = 3.0,
              labSize = 6.0,
              drawConnectors = TRUE)
```

```{r ,warning=FALSE, message=FALSE, results = 'hide'}
#| fig-width: 12
#| fig-height: 10

sce_metacells_alldatasets_perCT_10w$exN$DESeq2_lst$heatmap
```

# GSEA per celltype | 10w
```{r GSEA_perCelltype_metacells_alldatasets_COp1_10w,warning=FALSE, message=FALSE, results = 'hide', fig.height=12, fig.width=12}
gsea_10w = summarize_gsea_per_celltype(sce_metacells_alldatasets_perCT_10w, gs, params$gsea_n_pathways)

n_GOterms = list()
n_GOterms[["COp1_10w"]] = gsea_10w$goterm_summary
```

# Generate metacells + DEA | COp1 4m | per Celltype

```{r ,warning=FALSE, message=FALSE, results = 'hide'}
#| fig-width: 12
#| fig-height: 10

celltypes = unique(harmonized_seurat_COp1_4m$Celltype)

sce_metacells_alldatasets_perCT_4m = generate_metacells_dea_per_celltype(
  harmonized_seurat_COp1_4m, celltypes, params$amount_of_cells_to_aggregate, params$n_metacells_generate
)

saveRDS(sce_metacells_alldatasets_perCT_4m , file.path(object_storage_path , "metacell_sce_allCelltypes_COp1_4m.RDS"))

markers = sce_metacells_alldatasets_perCT_4m$exN$markers
EnhancedVolcano(markers,
              lab = rownames(markers),
              x = 'avg_log2FC',
              y = 'padj',
              title = paste0('Metacell COp1_4m ICE vs wt - ', "exN"),
              pCutoff = 0.05,
              FCcutoff = 1,
              pointSize = 3.0,
              labSize = 6.0,
              drawConnectors = TRUE)
```

```{r ,warning=FALSE, message=FALSE, results = 'hide'}
#| fig-width: 12
#| fig-height: 10

sce_metacells_alldatasets_perCT_4m$exN$DESeq2_lst$heatmap
```

# GSEA per celltype | COp1 4m
```{r GSEA_perCelltype_metacells_alldatasets_4m,warning=FALSE, message=FALSE, results = 'hide', fig.height=12, fig.width=12}
gsea_4m = summarize_gsea_per_celltype(sce_metacells_alldatasets_perCT_4m, gs, params$gsea_n_pathways)

n_GOterms[["COp1_4m"]] = gsea_4m$goterm_summary
```

# Generate metacells + DEA | COp2 | per Celltype

```{r ,warning=FALSE, message=FALSE, results = 'hide'}
#| fig-width: 12
#| fig-height: 10

celltypes = unique(harmonized_seurat_COp2_4dpi_4wpi$Celltype)

sce_metacells_alldatasets_perCT_COp2 = generate_metacells_dea_per_celltype(
  harmonized_seurat_COp2_4dpi_4wpi, celltypes, params$amount_of_cells_to_aggregate, params$n_metacells_generate
)

saveRDS(sce_metacells_alldatasets_perCT_COp2 , file.path(object_storage_path , "metacell_sce_allCelltypes_COp2.RDS"))

markers = sce_metacells_alldatasets_perCT_COp2$exN$markers
EnhancedVolcano(markers,
              lab = rownames(markers),
              x = 'avg_log2FC',
              y = 'padj',
              title = paste0('Metacell COp2 ICE vs wt - ', "exN"),
              pCutoff = 0.05,
              FCcutoff = 1,
              pointSize = 3.0,
              labSize = 6.0,
              drawConnectors = TRUE)
```

```{r ,warning=FALSE, message=FALSE, results = 'hide'}
#| fig-width: 12
#| fig-height: 10

sce_metacells_alldatasets_perCT_COp2$exN$DESeq2_lst$heatmap
```

# GSEA per celltype | COp2
```{r GSEA_perCelltype_metacells_alldatasets_COp2,warning=FALSE, message=FALSE, results = 'hide', fig.height=12, fig.width=12}
gsea_COp2 = summarize_gsea_per_celltype(sce_metacells_alldatasets_perCT_COp2, gs, params$gsea_n_pathways)

n_GOterms[["COp2"]] = gsea_COp2$goterm_summary
```

# plot relative amount of GOterms
```{r plot_nGOterms,warning=FALSE, message=FALSE, results = 'hide', fig.height=12, fig.width=10}
#| fig-width: 6
#| fig-height: 8

plot_data_clean <- bind_rows(n_GOterms, .id = "Dataset") %>%
  mutate(Dataset = factor(Dataset, levels = c("COp1_4m", "COp1_10w", "COp2")))

# 2. Generate the polished plot
ggplot(plot_data_clean, aes(x = Dataset, y = relative_n, fill = celltype)) +
  geom_col(width = 0.7, color = "white", linewidth = 0.2) + 
  scale_y_continuous(labels = scales::percent_format(), expand = c(0, 0)) + 
  scale_fill_viridis_d(option = "turbo", direction = 1) +
  labs(
    title = "Relative amount of GO terms",
    y = "Relative Proportion",
    fill = "Cell Type"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_line(color = "gray92"),
    
    plot.title = element_text(face = "bold", size = 16, margin = margin(b = 4)),
    plot.subtitle = element_text(color = "gray40", size = 11, margin = margin(b = 15)),
    axis.title.x = element_text(margin = margin(t = 12), face = "bold", size = 11),
    axis.title.y = element_text(margin = margin(r = 12), face = "bold", size = 11),
    
    legend.title = element_text(face = "bold", size = 11),
    legend.text = element_text(size = 10),
    legend.key.size = unit(0.5, "cm")
  )
```


# Save metacells | COp1_10w | IN/EN

**This is used for prediction**

```{r, warning=FALSE, message=FALSE, results = 'hide'}
saveRDS(sce_metacells_alldatasets_perCT_10w$iN1$sce, file.path(object_storage_path , "metacell_sce_IN_organoids_COp1_10w.RDS"))
saveRDS(sce_metacells_alldatasets_perCT_10w$exN$sce, file.path(object_storage_path , "metacell_sce_EN_organoids_COp1_10w.RDS"))
```

**Here we are trying to understand if phenotype is a variable that can explain the variability in the data**

::: panel-tabset

## DEA IN

```{r, warning=FALSE, message=FALSE, results = 'hide', fig.height=8, fig.width=8}
markers = plot_dea_tab(
  sce_metacells_alldatasets_perCT_10w$iN1$DESeq2_lst,
  file.path(DEA_results_path , "DEA_metacell_ICE_vs_wt_iN_COp1_10w_organoids.csv")
)
```

## DEA EN

```{r, warning=FALSE, message=FALSE, results = 'hide', fig.height=8, fig.width=8}
markers = plot_dea_tab(
  sce_metacells_alldatasets_perCT_10w$exN$DESeq2_lst,
  file.path(DEA_results_path , "DEA_metacell_ICE_vs_wt_exN_COp1_10w_organoids.csv")
)
```

## GSEA IN

```{r DEA_GSEA_IN_metacell_10w, fig.height=12, fig.width=12}
x = run_gsea_tab(
  markers_csv_path = file.path(DEA_results_path , "DEA_metacell_ICE_vs_wt_iN_COp1_10w_organoids.csv"),
  gs = gs, n_pathways = params$gsea_n_pathways,
  up_txt_path = file.path(gsea_results_path , "GOterms_metacell_ICE_vs_wt_COp1_10w_iN_upregulated_ICE_organoids.txt"),
  down_txt_path = file.path(gsea_results_path , "GOterms_metacell_ICE_vs_wt_COp1_10w_iN_downregulated_ICE_organoids.txt"),
  title_label = "metacell COp1_10w_IN - ICE vs wt"
)
```

## GSEA EN

```{r DEA_GSEA_EN_metacell_10w, fig.height=12, fig.width=12}
x = run_gsea_tab(
  markers_csv_path = file.path(DEA_results_path , "DEA_metacell_ICE_vs_wt_exN_COp1_10w_organoids.csv"),
  gs = gs, n_pathways = params$gsea_n_pathways,
  up_txt_path = file.path(gsea_results_path , "GOterms_metacell_ICE_vs_wt_COp1_10w_exN_upregulated_ICE_organoids.txt"),
  down_txt_path = file.path(gsea_results_path , "GOterms_metacell_ICE_vs_wt_COp1_10w_exN_downregulated_ICE_organoids.txt"),
  title_label = "metacell COp1_10w_EN - ICE vs wt"
)
```
:::

# Save metacells | COp1_4m | IN/EN

**This is used for prediction**

```{r, warning=FALSE, message=FALSE, results = 'hide'}
saveRDS(sce_metacells_alldatasets_perCT_4m$iN1$sce, file.path(object_storage_path , "metacell_sce_IN_organoids_COp1_4m.RDS"))
saveRDS(sce_metacells_alldatasets_perCT_4m$exN$sce, file.path(object_storage_path , "metacell_sce_EN_organoids_COp1_4m.RDS"))
```


::: panel-tabset

## DEA IN

```{r, warning=FALSE, message=FALSE, results = 'hide', fig.height=8, fig.width=8}
markers = plot_dea_tab(
  sce_metacells_alldatasets_perCT_4m$iN1$DESeq2_lst,
  file.path(DEA_results_path , "DEA_metacell_ICE_vs_wt_iN_COp1_4m_organoids.csv")
)
```

## DEA EN

```{r, warning=FALSE, message=FALSE, results = 'hide', fig.height=8, fig.width=8}
markers = plot_dea_tab(
  sce_metacells_alldatasets_perCT_4m$exN$DESeq2_lst,
  file.path(DEA_results_path , "DEA_metacell_ICE_vs_wt_exN_COp1_4m_organoids.csv")
)
```

## GSEA IN

```{r DEA_GSEA_IN_metacell_4m, fig.height=12, fig.width=12}
x = run_gsea_tab(
  markers_csv_path = file.path(DEA_results_path , "DEA_metacell_ICE_vs_wt_iN_COp1_4m_organoids.csv"),
  gs = gs, n_pathways = params$gsea_n_pathways,
  up_txt_path = file.path(gsea_results_path , "GOterms_metacell_ICE_vs_wt_COp1_4m_iN_upregulated_ICE_organoids.txt"),
  down_txt_path = file.path(gsea_results_path , "GOterms_metacell_ICE_vs_wt_COp1_4m_iN_downregulated_ICE_organoids.txt"),
  title_label = "metacell COp1_4m_IN - ICE vs wt"
)
```

## GSEA EN

```{r DEA_GSEA_EN_metacell_4m, fig.height=12, fig.width=12}
x = run_gsea_tab(
  markers_csv_path = file.path(DEA_results_path , "DEA_metacell_ICE_vs_wt_exN_COp1_4m_organoids.csv"),
  gs = gs, n_pathways = params$gsea_n_pathways,
  up_txt_path = file.path(gsea_results_path , "GOterms_metacell_ICE_vs_wt_COp1_4m_exN_upregulated_ICE_organoids.txt"),
  down_txt_path = file.path(gsea_results_path , "GOterms_metacell_ICE_vs_wt_COp1_4m_exN_downregulated_ICE_organoids.txt"),
  title_label = "metacell COp1_4m_EN - ICE vs wt"
)
```
:::


# Generate metacell | COp2 4dpi | IN/EN

This is used for prediction

```{r,warning=FALSE, message=FALSE, results = 'hide'}
sce_metacells_COp2_4dpi = generate_metacell_by_timepoint(
  harmonized_seurat_COp2_4dpi_4wpi, c("iN1", "exN"), "4dpi",
  params$amount_of_cells_to_aggregate, params$n_metacells_generate,
  file.path(DEA_results_path, "DEA_metacell_ICE_vs_wt")
)

saveRDS(sce_metacells_COp2_4dpi$iN1, file.path(object_storage_path , "metacell_sce_IN_organoids_COp2_4dpi.RDS"))
saveRDS(sce_metacells_COp2_4dpi$exN, file.path(object_storage_path , "metacell_sce_EN_organoids_COp2_4dpi.RDS"))
```

# Generate metacell | COp2 4wpi | IN/EN

This is used for prediction

```{r,warning=FALSE, message=FALSE, results = 'hide'}
sce_metacells_COp2_4wpi = generate_metacell_by_timepoint(
  harmonized_seurat_COp2_4dpi_4wpi, c("iN1", "exN"), "4wpi",
  params$amount_of_cells_to_aggregate, params$n_metacells_generate,
  file.path(DEA_results_path, "DEA_metacell_ICE_vs_wt")
)

saveRDS(sce_metacells_COp2_4wpi$iN1, file.path(object_storage_path , "metacell_sce_IN_organoids_COp2_4wpi.RDS"))
saveRDS(sce_metacells_COp2_4wpi$exN, file.path(object_storage_path , "metacell_sce_EN_organoids_COp2_4wpi.RDS"))
```


# DEA | all celltypes + all datasets together

```{r DEA_alldatasets_allcelltypes_together,warning=FALSE, message=FALSE, results = 'hide', fig.height=12, fig.width=12}
all_datasets = list(
  "10w" = sce_metacells_alldatasets_perCT_10w,
  "4m"  = sce_metacells_alldatasets_perCT_4m,
  "COp2" = sce_metacells_alldatasets_perCT_COp2
)

all_celltypes = unique(unlist(lapply(all_datasets, names)))

# For each celltype, pull its sce from every dataset it appears in, tag which dataset
# it came from, and merge those together on their shared genes.
merged_list_by_celltype = lapply(all_celltypes, function(ct) {
  sce_elements = lapply(names(all_datasets), function(ds_name) {
    sce_obj = all_datasets[[ds_name]][[ct]]$sce
    if (is.null(sce_obj)) return(NULL)
    colData(sce_obj)$dataset_batch = ds_name
    sce_obj
  })
  sce_elements = Filter(Negate(is.null), sce_elements)
  merge_sces_common_genes(sce_elements)
}) %>% setNames(all_celltypes)

# Merge every celltype's combined object together, again restricted to shared genes
master_merged_sce = merge_sces_common_genes(merged_list_by_celltype)

DESeq2_lst = metacell_DESeq2_DEA(sce = master_merged_sce, phenotype_column = "phenotype", dataset_column = "dataset", design_formula = as.formula("~ maturation_stage + Protocol + Celltype + phenotype"), title_corHeatmap = "metacell correlation across all datasets and all celltypes", contrast = c("phenotype", "ICE", "wt"), organism = "organoid")

markers = DESeq2_lst$markers %>% na.omit()

EnhancedVolcano(markers,
              lab = rownames(markers),
              x = 'avg_log2FC',
              y = 'padj',
              title = 'Metacell ICE vs wt - all celltypes and datasets together',
              pCutoff = 0.05,
              FCcutoff = 1,
              pointSize = 3.0,
              labSize = 6.0,
              drawConnectors = TRUE)
```

# DEA | neurons + all datasets together

```{r DEA_alldatasets_neurons_together,warning=FALSE, message=FALSE, results = 'hide', fig.height=12, fig.width=12}
all_datasets_neurons = list(
  "EN_10w" = sce_metacells_alldatasets_perCT_10w$exN$sce, 
  "IN_10w" = sce_metacells_alldatasets_perCT_10w$iN1$sce,
  "EN_4m" = sce_metacells_alldatasets_perCT_4m$exN$sce, 
  "IN_4m" = sce_metacells_alldatasets_perCT_4m$iN1$sce,
  "EN_COp2" = sce_metacells_alldatasets_perCT_COp2$exN$sce, 
  "IN_COp2" = sce_metacells_alldatasets_perCT_COp2$iN1$sce)

master_merged_sce = merge_sces_common_genes(all_datasets_neurons)

DESeq2_lst = metacell_DESeq2_DEA(sce = master_merged_sce, phenotype_column = "phenotype", dataset_column = "dataset", design_formula = as.formula("~ maturation_stage + Protocol + Celltype + phenotype"), title_corHeatmap = "metacell correlation across all datasets but only neurons", contrast = c("phenotype", "ICE", "wt"), organism = "organoid")

markers = DESeq2_lst$markers %>% na.omit()

write.csv(markers, file.path(DEA_results_path , "DEA_metacell_ICE_vs_wt_onlyneurons_allDatasets_organoids.csv"))

EnhancedVolcano(markers,
              lab = rownames(markers),
              x = 'avg_log2FC',
              y = 'padj',
              title = 'Metacell ICE vs wt - datasets together but only neurons',
              pCutoff = 0.05,
              FCcutoff = 1,
              pointSize = 3.0,
              labSize = 6.0,
              drawConnectors = TRUE)
```

# GSEA | neurons + all datasets together

```{r GSEA_alldatasets_neurons_together,warning=FALSE, message=FALSE, results = 'hide', fig.height=12, fig.width=12}
gsea = gsea_analysis(markers = markers, cluster = "trending clusters", gs = gs, n_pathways = params$gsea_n_pathways, plot_type = "bar")

write.table(gsea$gsea_df %>% filter(NES > 0) %>% select(c("hsapiens_BP_ID", "p.adjust")),  file.path(gsea_results_path , "GOterms_metacell_ICE_vs_wt_neurons_alldatasets_upregulated_ICE_organoids.txt"), row.names = F, quote = F)
write.table(gsea$gsea_df %>% filter(NES < 0) %>% select(c("hsapiens_BP_ID", "p.adjust")),  file.path(gsea_results_path , "GOterms_metacell_ICE_vs_wt_neurons_alldatasets_downregulated_ICE_organoids.txt"), row.names = F, quote = F)

# Generate treemap of NES > 0 and NES < 0
x = plot_ReducedGOterms(gsea_df = gsea$gsea_df, title_label = "metacell neurons | all datasets | - ICE vs wt")
```


# Generate metacell | All datasets | All Celltypes

```{r,warning=FALSE, message=FALSE, results = 'hide'}
set.seed(1)
# NOTE: only 2 of the underlying organoid datasets are at 4wpi (one ICE, one wt) vs
# 10 at 4dpi
celltypes_timepoint_comparison = unique(harmonized_seurat$Celltype)

metacells_timepoint_comparison = list()
for (ct in celltypes_timepoint_comparison) {
  message(ct)
  sce = generate_metacell_celltype_timepointComparison(
    harmonized_seurat, ct, params$amount_of_cells_to_aggregate, params$n_metacells_generate
  )
  if (length(unique(sce$timepoint)) == 2) {
    metacells_timepoint_comparison[[ct]] = sce
  } else {
    message(ct, " - only one timepoint present, excluded")
  }
}

saveRDS(metacells_timepoint_comparison, file.path(object_storage_path, "metacell_sce_perCelltype_timepointComparison_organoids.RDS"))
```

# DEA + GSEA from chunk above | All celltypes


| Table / test | Question it answers |
|---|---|
| `markers_interaction_allCT` (LRT) | Does the size of the ICE effect differ significantly between 4dpi and 4wpi? (padj per gene) |
| `phenotype_at_4dpi_allCT` (Wald) | Is this gene higher in ICE vs wt, at 4dpi only? (direction + magnitude) |
| `phenotype_at_4wpi_allCT` (Wald) | Same question, at 4wpi only |
| `effect_change` (4wpi minus 4dpi) | How did the ICE effect change between timepoints - grew, shrank, or reversed? |

```{r,warning=FALSE, message=FALSE, results = 'hide'}
design_full_allCT = as.formula("~ Celltype + timepoint * phenotype")
design_reduced_allCT = as.formula("~ Celltype + timepoint + phenotype")

merged_sce_timepoint_allCT = merge_sces_common_genes(metacells_timepoint_comparison)
merged_sce_timepoint_allCT$Celltype = factor(merged_sce_timepoint_allCT$Celltype)

# Does the ICE-vs-wt effect differ between 4dpi and 4wpi? (all celltypes pooled, controlling for Celltype)
result_allCT = run_timepoint_phenotype_interaction_dea(merged_sce_timepoint_allCT, design_full_allCT, design_reduced_allCT)
markers_interaction_allCT = result_allCT$markers
write.csv(markers_interaction_allCT, file.path(DEA_results_path, "DEA_LRT_timepoint_ICE_allCelltypes_organoids.csv"))

EnhancedVolcano(markers_interaction_allCT,
              lab = rownames(markers_interaction_allCT),
              x = 'avg_log2FC', y = 'padj',
              title = 'Interaction: timepoint(4wpi vs 4dpi) x ICE vs wt - all celltypes',
              pCutoff = 0.05, FCcutoff = 1, pointSize = 3.0, labSize = 6.0, drawConnectors = TRUE)

design_full_allCT = as.formula("~ Celltype + timepoint * phenotype")

# ICE vs wt effect, separately at each timepoint
phenotype_at_4dpi_allCT = get_phenotype_effect_at_timepoint(merged_sce_timepoint_allCT, "4dpi", design_full_allCT)
write.csv(phenotype_at_4dpi_allCT, file.path(DEA_results_path, "DEA_ICE_vs_wt_at4dpi_allCelltypes_organoids.csv"))

phenotype_at_4wpi_allCT = get_phenotype_effect_at_timepoint(merged_sce_timepoint_allCT, "4wpi", design_full_allCT)
write.csv(phenotype_at_4wpi_allCT, file.path(DEA_results_path, "DEA_ICE_vs_wt_at4wpi_allCelltypes_organoids.csv"))

# Combined interpretation table: interaction significance + effect at each timepoint +
# direction of change. Answers "does gene X have
# higher ICE expression at 4wpi compared to 4dpi"
interpretation_table_allCT = phenotype_at_4dpi_allCT %>%
  select(gene, log2FC_at_4dpi = avg_log2FC, padj_at_4dpi = padj) %>%
  inner_join(
    phenotype_at_4wpi_allCT %>% select(gene, log2FC_at_4wpi = avg_log2FC, padj_at_4wpi = padj),
    by = "gene"
  ) %>%
  inner_join(markers_interaction_allCT %>% select(gene, padj_interaction = padj), by = "gene") %>%
  mutate(effect_change = log2FC_at_4wpi - log2FC_at_4dpi) %>%
  arrange(padj_interaction)%>%
  mutate(
    direction_at_4dpi = case_when(log2FC_at_4dpi > 0 ~ "ICE > wt", log2FC_at_4dpi < 0 ~ "wt > ICE", TRUE ~ "equal"),
    direction_at_4wpi = case_when(log2FC_at_4wpi > 0 ~ "ICE > wt", log2FC_at_4wpi < 0 ~ "wt > ICE", TRUE ~ "equal"),
    pattern = case_when(
      sign(log2FC_at_4dpi) != sign(log2FC_at_4wpi) & sign(log2FC_at_4dpi) != 0 & sign(log2FC_at_4wpi) != 0 ~ "reversal",
      TRUE ~ "same direction, magnitude changed"
    )
  )

genes_appearing_at_4wpi_only = interpretation_table_allCT %>%
  filter(
    padj_interaction < 0.05,           # the change over time is statistically siginifcant
    padj_at_4dpi > 0.05 | abs(log2FC_at_4dpi) < 0.5,   # ~absent at 4dpi
    padj_at_4wpi < 0.05,               # present (significant) at 4wpi
  ) %>%
  arrange(desc(abs(effect_change)))

write.csv(interpretation_table_allCT, file.path(DEA_results_path, "DEA_ICE_effect_change_4dpi_vs_4wpi_allCelltypes_organoids.csv"))
write.csv(genes_appearing_at_4wpi_only, file.path(DEA_results_path, "DEA_ICE_effect_change_at_4wpi_only_allCelltypes_organoids.csv"))

# GSEA for genes/pathways specifically emerging at 4wpi - all celltypes together
if (nrow(genes_appearing_at_4wpi_only) > 0) {
  gsea_input_allCT_emerging = genes_appearing_at_4wpi_only %>%
    dplyr::rename(avg_log2FC = log2FC_at_4wpi) %>%
    select(gene, avg_log2FC)

  gsea_emerging_4wpi_allCT = gsea_analysis(
    markers = gsea_input_allCT_emerging, cluster = "trending clusters", gs = gs,
    n_pathways = params$gsea_n_pathways, plot_type = "bar"
  )

  if (length(gsea_emerging_4wpi_allCT) > 1) {
    write.table(gsea_emerging_4wpi_allCT$gsea_df %>% filter(NES > 0) %>% select(c("hsapiens_BP_ID", "Description", "p.adjust")),
                file.path(gsea_results_path, "GOterms_emerging4wpi_allCelltypes_upregulatedICE_organoids.txt"), row.names = FALSE, quote = FALSE)
    write.table(gsea_emerging_4wpi_allCT$gsea_df %>% filter(NES < 0) %>% select(c("hsapiens_BP_ID", "Description", "p.adjust")),
                file.path(gsea_results_path, "GOterms_emerging4wpi_allCelltypes_downregulatedICE_organoids.txt"), row.names = FALSE, quote = FALSE)

    x = plot_ReducedGOterms(gsea_df = gsea_emerging_4wpi_allCT$gsea_df, title_label = "all celltypes | genes emerging at 4wpi")
  } else {
    message("No enriched GOterms for genes emerging at 4wpi - all celltypes")
  }
} else {
  message("No genes passed the 'emerging at 4wpi' filter - all celltypes")
}
```

# DEA + GSEA from chunk above | Per celltype

```{r,warning=FALSE, message=FALSE, results = 'hide'}
design_full_perCT = as.formula("~ timepoint * phenotype")

phenotype_at_4dpi_perCT = list()
phenotype_at_4wpi_perCT = list()
interpretation_table_perCT_list = list()
genes_appearing_at_4wpi_perCT = list()
gsea_emerging_4wpi_perCT = list()

for (ct in names(metacells_timepoint_comparison)) {
  cat("\n\n## ", ct, "\n\n")

  sce = metacells_timepoint_comparison[[ct]]

  # --- DEA: ICE vs wt, separately at each timepoint ---
  phenotype_at_4dpi_perCT[[ct]] = get_phenotype_effect_at_timepoint(sce, "4dpi", design_full_perCT)
  write.csv(phenotype_at_4dpi_perCT[[ct]], file.path(DEA_results_path, paste0("DEA_ICE_vs_wt_at4dpi_", ct, "_organoids.csv")))

  phenotype_at_4wpi_perCT[[ct]] = get_phenotype_effect_at_timepoint(sce, "4wpi", design_full_perCT)
  write.csv(phenotype_at_4wpi_perCT[[ct]], file.path(DEA_results_path, paste0("DEA_ICE_vs_wt_at4wpi_", ct, "_organoids.csv")))

  # --- Combined table: effect at each timepoint ---
  interpretation_table_perCT = phenotype_at_4dpi_perCT[[ct]] %>%
    select(gene, log2FC_at_4dpi = avg_log2FC, padj_at_4dpi = padj) %>%
    inner_join(
      phenotype_at_4wpi_perCT[[ct]] %>% select(gene, log2FC_at_4wpi = avg_log2FC, padj_at_4wpi = padj),
      by = "gene"
    ) %>%
    mutate(effect_change = log2FC_at_4wpi - log2FC_at_4dpi) %>%
    arrange(desc(abs(effect_change))) %>%
    mutate(
      direction_at_4dpi = case_when(log2FC_at_4dpi > 0 ~ "ICE > wt", log2FC_at_4dpi < 0 ~ "wt > ICE", TRUE ~ "equal"),
      direction_at_4wpi = case_when(log2FC_at_4wpi > 0 ~ "ICE > wt", log2FC_at_4wpi < 0 ~ "wt > ICE", TRUE ~ "equal"),
      pattern = case_when(
        sign(log2FC_at_4dpi) != sign(log2FC_at_4wpi) & sign(log2FC_at_4dpi) != 0 & sign(log2FC_at_4wpi) != 0 ~ "reversal",
        TRUE ~ "same direction, magnitude changed"
      )
    )

  interpretation_table_perCT_list[[ct]] = interpretation_table_perCT
  #write.csv(interpretation_table_perCT, file.path(DEA_results_path, paste0("DEA_ICE_effect_change_4dpi_vs_4wpi_", ct, "_organoids.csv")))

  # --- Genes with an ICE-vs-wt difference specifically at 4wpi, essentially absent at 4dpi ---
  genes_ct = interpretation_table_perCT %>%
    filter(
      padj_at_4dpi > 0.05 | abs(log2FC_at_4dpi) < 0.5,   # ~absent at 4dpi
      padj_at_4wpi < 0.05,                               # present (significant) at 4wpi
    ) %>%
    arrange(desc(abs(effect_change)))

  genes_appearing_at_4wpi_perCT[[ct]] = genes_ct
  #write.csv(genes_ct, file.path(DEA_results_path, paste0("DEA_ICE_genes_emerging_at4wpi_", ct, "_organoids.csv")))

  # --- GSEA on the genes emerging at 4wpi ---
  if (nrow(genes_ct) == 0) {
    message(ct, " - no genes passed the 'emerging at 4wpi' filter")
    next
  }

  gsea_input_ct = genes_ct %>%
    dplyr::rename(avg_log2FC = log2FC_at_4wpi) %>%
    select(gene, avg_log2FC)

  gsea = gsea_analysis(markers = gsea_input_ct, cluster = ct, gs = gs, n_pathways = params$gsea_n_pathways, plot_type = "none")

  if (length(gsea) == 1) {
    message(ct, " has no enriched GOterms for genes emerging at 4wpi")
    next
  }

  lst = plot_ReducedGOterms(gsea_df = gsea$gsea_df, title_label = paste0(ct, " - genes emerging at 4wpi"))
  gsea_emerging_4wpi_perCT[[ct]] = lst
}

saveRDS(interpretation_table_perCT_list, file.path(object_storage_path, "DEA_ICE_effect_change_perCelltype_organoids.RDS"))
saveRDS(genes_appearing_at_4wpi_perCT, file.path(object_storage_path, "DEA_ICE_genes_emerging_at4wpi_perCelltype_organoids.RDS"))
saveRDS(gsea_emerging_4wpi_perCT, file.path(gsea_results_path, "gsea_emerging_4wpi_perCelltype_organoids.RDS"))
```

# sessioninfo
```{r}
sessionInfo()
```
